Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-10-02 paper/paper.tex
edits to paper in mannik section Tested everything on quipu.

Huang's simulations use total MinD and MinE concentrations of
$1,000/\micron$ and $350/\micron$, respectively, in a cylindrical cell of
radius $0.5\micron$, and in our (non-cylindrical) cells we use the same
number of proteins per unit volume.  These concentration values are
$1273\micron^{-3}$ and $446\micron^{-3}$, respectively. \fixme{It's been
  a while so might want to check this to make sure it's true.}  We
construct a 3D grid in cartesian coordinates, with a grid spacing of
.05\micron.

@@

\fixme{WHERE TO SAY??? : I think we should describe the ``starry
  night'' effect in the stochastic model.}

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{../paper/plot-ave}
  \caption{We display here arrows depicting successive maxima in space
    and and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is \fixme{340} seconds, which is the same
    period of time depicted in the experimental data plots of Mannik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild type pill
    shape, we only cover \fixme{???} seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by Mannik of the MinD maxima behavoir
    and the bottom two rows show our simulations using the stochastic
    and deterministic models, respectively.  We simulated
    approximations to the two shapes observed by Mannik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

@@

As describe in Section~\ref{sec:model-method-shapes}, we will focus on
simulation of four illustrative flattened cell shapes: two shapes
created to replicate those published by
Mannik~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
cellular shape irregularity and assymetry, and that of overall size
and aspect ratio.  We note that shape \emph{B} is more irregular than
shape \emph{A}, and in particular features a prominant region of
concavity on its left-hand side.  For each of these shapes (in
addition to the wild type pill shape discussed previously) we have
simulated in excess of \fixme{???}  seconds of evolution of the MinD
system using the deterministic version of Huang's
model~\cite{huang2003dynamic}, as well as the stochastic model.

From each of these flattened simulations, we have chosen a typical 350
second segment to compare with the results of Mannik exemplified in
arrow plots in which the arrow heads show the location of sequential
MinD maxima within the cell (over the same time
interval)~\cite{mannik2012robustness}.  In addition to arrows between
successive maxima in space and time, we plot as a colored background
the density of MinD protiens averaged over the same time period.  We
note that we have manually verified that our (computer-generated)
arrow plots also reflect a human interpretation of a movie of the same
data.  Finally, for comparison we present the same plot for a
wild-type cell, with a time period of \fixme{???}  seconds to account
for its short period of oscillation.

@@

The predictions of the stochastic model as seen in
Fig.~\ref{randst-plot-ave} are roughly similar to the experimental
results in terms of irregularity of the locations of maxima, although
there are locations on the edges of the cells where experiment shows
maxima occuring that we never see in our simulations, suggesting that
the model does not precisely reflect the experiment.  We also note
that the stadium shapes in Fig.~\ref{randst-plot-ave} appear
qualitatively similar to the irregular shapes in terms of the
locations of their maxima.

However, Fig.~\ref{randst-plot-ave} leaves unclear the importance of
irregularity versus overall thickness and aspect ratio in developing
an understanding of the origin of the experimentally observed
behavior.  Although in the movies the stadiums appear to display a
more regular bipolar oscillation, the arrow plots do not provide a
convincing demonstration.  To illustrate the difference in behavior
between shape \emph{A} and stadium \emph{A}, we return in
Fig.~\ref{fig:corr-pancake-A} to the correlation function between the
number of proteins in a segment at the top of the cell and the number
of proteins in a segment at the bottom of the cell.  We computed these
correlation functions using over \fixme{???} seconds of simulation.
We found with the stochastic model that the experimental shape
\emph{A} has a coherence time of under 4 periods, while the
corresponding stadium \emph{A} has a coherence time of 10 periods.  In
each case, the deterministic model is perfectly periodic. This
confirms that with the stochastic model the stadium shape does indeed
have a far more regular bipolar oscillation than the irregular shape
\emph{A}.  When we compare shape \emph{B} and stadium \emph{B}, we see
a similar effect, although both coherence times are somewhat smaller
(2.7 and 6.2, respectively).  From these results we conclude that the
irregularity seen experimentally is indeed exacerbated by the
irregular shapes taken on by the cells forced into a thin slit.

\section{Conclusion}
We have simulated flattened cell shapes similar to those observed
experimentally by Mannik \emph{et al.}.  We find that the
deterministic model~\fixme{cite} predicts behavior that completely
contradicts what is seen experimentally: it predicts stable and
periodic bipolar oscillation, with a minor intermediate maximum
appearing in shape \emph{B}.  Thus we conclude that this method is
inconsistent with experiment, and should not be relied upon for
predictions of the behavior of the MinD system.

As found in previous studies~\fixme{cite}, we observe that the
stochastic model predicts behavior in wild type cells that is
consistent both with experiment and with the deterministic model.  In
contrast, the stochastic model~\fixme{cite} predicts spatially
irregular formation of maxima for all the flattened cells that we
studied.  In order to better understand the irregular oscillations
observed in experiment, we simulated flattened ``stadium'' shapes in
order to distinguish between the effect of flattening and the effect
of irregular shape.  We found that although the stadium cells
exhibited somewhat more irregular locations of maxima than the
wild-type cells, their oscillation was far more periodic and bipolar
than the irregular cell shapes observed experimentally.  Thus we
confirm that the disruption of the MinD oscillations seen by Mannik
\emph{et al.} is largely due to the irregularity of cell shapes
observed, as opposed to the size or aspect ratio.

@@

>>>>

Huang's simulations use total MinD and MinE concentrations of
$1,000/\micron$ and $350/\micron$, respectively, in a cylindrical cell
of radius $0.5\micron$, and in our (non-cylindrical) cells we use the
same number of proteins per unit volume.  These concentration values
are $1273\micron^{-3}$ and $446\micron^{-3}$, respectively. We
construct a 3D grid in cartesian coordinates, with a grid spacing of
.05\micron.

@@

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{../paper/plot-ave}
  \caption{We display here arrows depicting successive maxima in space
    and and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is 350 seconds, which is the same
    period of time depicted in the experimental data plots of Mannik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild type pill
    shape, we only cover 250 seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by Mannik of the MinD maxima behavoir
    and the bottom two rows show our simulations using the stochastic
    and deterministic models, respectively.  We simulated
    approximations to the two shapes observed by Mannik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

@@

As describe in Section~\ref{sec:model-method-shapes}, we will focus on
simulation of four illustrative flattened cell shapes: two shapes
created to replicate those published by
Mannik~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
cellular shape irregularity and assymetry, and that of overall size
and aspect ratio.  We note that shape \emph{B} is more irregular than
shape \emph{A}, and in particular features a prominant region of
concavity on its left-hand side.  For each of these shapes (in
addition to the wild type pill shape discussed previously) we have
simulated in excess of 15000 seconds of evolution of the MinD
system using the deterministic version of Huang's
model~\cite{huang2003dynamic}, as well as the stochastic model.

From each of these flattened simulations, we have chosen a typical 350
second segment to compare with the puplished results of
Mannik~\cite{mannik2012robustness}, exemplified in arrow plots in
which the arrow heads show the location of sequential MinD maxima
within the cell (in Fig.~\ref{randst-plot-ave}).  In addition to
arrows between successive maxima in space and time, we plot as a
colored background the density of MinD protiens averaged over the same
time period.  We note that we have manually verified that our
(computer-generated) arrow plots also reflect a human interpretation
of a movie of the same data.  Finally, for comparison we present the
same plot for a wild-type cell, with a time period of 250 seconds to
account for its short period of oscillation.

@@

The predictions of the stochastic model are roughly similar to the
experimental results in terms of irregularity of the locations of
maxima, although there are locations on the edges of the cells where
experiment shows maxima occuring that we never see in our simulations,
suggesting that the model does not precisely reflect the experiment.
We also note that the stadium shapes in Fig.~\ref{randst-plot-ave}
appear to have maxima location irregularity that is qualitatively
similar to our mannik shapes and to experiment.

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
irregularity versus overall thickness and aspect ratio in developing
an understanding of the origin of the experimentally observed
behavior.  Although in the movies the stadiums appear to display, on
average over long time scales, a more regular bipolar oscillation than
the irregular mannik shapes, the limited time range of the arrow plots
(350 seconds) is inadequate when studying this difference in behavoir.
We therefore return to the correlation function between the number of
proteins in a segment at the top of the cell and the number of
proteins in a segment at the bottom of the cell.  For each shape we
computed these correlation functions using the entirety of our over
15000 seconds of simulation data. Fig.~\ref{fig:corr-pancake-A} shows
this plot for mannik \emph{A} and stadium \emph{A}.  E. Coli cell
doubling times vary widely under alternate conditions, but will often
be between 20 minutes to 100 minutes~\cite{pierucci1972chromosome},
while flourescent microscopy experiments have shown that the FtsZ ring
builds itself with a half life of roughly 30
seconds~\cite{stricker2002rapid}.  Here we plot correlations of
correlation times that range from 0 up to 600, a biologically relevent
timescale.

We see that both in the mannik \emph{A} and stadium \emph{A}, the
deterministic model is perfectly periodic.  The vary slight drop in
correlation visible in the deterministic data is most likely due to
the fact that even with our large amount of simulation time, the end
of the simulation (sudden cutoff) will still have an effect. Both the
deterministic and stochastic model correlations have the same number
of proteins in their cells, and have been normalized with the same
normalization factor, so that they can be directly compared against
eachother.  At $\tau=0$, the stochastic correlation is already seen to
be lower than the deterministic, reflecting the fact that the
stochastic data now and then shows maxima or minima at both ends
simultaneously.  the stochastic model that the experimental shape
\emph{A} has a coherence time of 4.7 periods, while the corresponding
stadium \emph{A} has a coherence time of just over 6. This difference
in coherence time is not as extreme as we initially guessed, and it
shows that the regularity in the cells is not overly sensitive to the
regularity and symmetry of the shape.  It seems that it is the size
and aspect ratio that are more important in dictating the regularity
of bipolar oscillation.

\section{Conclusion}
We have simulated flattened cell shapes similar to those observed
experimentally by Mannik \emph{et al.}.  We find that the
deterministic model~\cite{huang2003dynamic} predicts behavior that
completely contradicts what is seen experimentally: it predicts stable
and periodic bipolar oscillation, with a minor intermediate maximum
appearing in shape \emph{B}.  Thus we conclude that this method is
inconsistent with experiment, and should not be relied upon for
predictions of the behavior of the MinD system.

As found in previous studies~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, we observe that the stochastic model
predicts qualitative behavior in wild type cells that is broadly
consistent both with experiment and with the deterministic
model. However, even in these simple pill-shaped cells the stochastic
model shows more variation in the location of MinD maxima on either
end of the cell. This model also predicts spatially irregular
formation of maxima for all the flattened cells that we studied.  In
order to better understand the irregular oscillations observed in
experiment, we simulated flattened ``stadium'' shapes in order to
distinguish between the effect of flattening and the effect of
irregular shape.  We did not find a significant difference in the
coherence of bipolar oscillations within these stadium shapes and the
irregular shapes published by Mannik \emph{et al.}.  Thus we conclude
that the disruption of the MinD oscillations seen by Mannik \emph{et
  al.}  is largely due to the aspect ratio of the cell shape, as
opposed to its irregularity.

@@

%% mannik B 11500 seconds, mannik A 15600 seconds, stadium A 19120  seconds, pill 19990 seconds
%% = simulation time that deterministic correlation data covers.
